---
title: '"Data Analysis in R Part 3: Data Modeling and Machine Learning"'
author: "Stephen Guest"
date: "November 8, 2016"
output: html_document
---

First things first, let's load in the caret library so that it will be available for us to use.

```{r}
library(caret)
set.seed(1364)
```

Now we'll load in the Alzheimers data set that we've been working with in previous sessions.

```{r}
Alz <- read.csv(file.choose())
dim(Alz)
summary(Alz$gender)
```

Before we go on let's clean up one of our variables that has some inconsistencies

```{r}
Alz$gender <- ifelse(Alz$gender == 'female', 'Female', ifelse(Alz$gender == 'M', 'Male', ifelse(Alz$gender == 'male', 'Male', ifelse(Alz$gender == 'Female', 'Female', 'Male'))))
Alz$gender <- as.factor(Alz$gender)
```

**Partitioning Data Into Training and Test Sets**
Typically one of the first things we want to do before building any prediction model is to take our data and split it up into a 'training' set that will be used to teach the model and a 'test' set that will be used as part of validating the model

The caret package has some really simple built in functions for partitioning data.
Let's use **createDataPartition()** to split the Alzheimers data set into a train and test set

```{r}
samples_for_training <- createDataPartition(Alz$response, p=.75, list=F)
training_samples <- Alz[samples_for_training,]
test_samples <- Alz[-samples_for_training,]
```

**Ready to Build a Model**
We're now ready to build a model using any of the machine learning algorithms that are built into caret

There are >200 machine learning algorithms available in the caret package

Whichever algorithm we choose we can generate a model using **train()** function

```{r, message=FALSE, warning=FALSE, results="hide"}
my_model_gbm <- train(response ~ ., data = training_samples, method = "gbm")
```

That's it.  
We've got a model now that we can use to make predictions.

Taking a look at the model we built

```{r}
my_model_gbm
```

**Generating Predictions with our model**
Let's make some predictions using the test data set and see how the model performs

To generate predictions we use the **predict()** function

```{r}
my_gbm_predictions <- predict(my_model_gbm, newdata = test_samples)
```

**Model Performance**
Visualizing model performance with a confusion matrix!

```{r}
cMatrix_gbm <- confusionMatrix(my_gbm_predictions, test_samples$response)
cMatrix_gbm
```

**Switching Algorithms**
Caret makes it exceptionally easy to try a different algorithm

```{r, message=FALSE, warning=FALSE}
my_model_glm <- train(response ~ ., data = training_samples, method = "glm")
my_glm_predictions <- predict(my_model_glm, newdata = test_samples)
cMatrix_glm <- confusionMatrix(my_glm_predictions, test_samples$response)
cMatrix_glm
```

**compare the two models**

```{r}
cMatrix_gbm$overall[1]
cMatrix_glm$overall[1]
```

**Imporoving our models**
For some algorithms to run or be optimized we might need to preprocess data

```{r}
hist(Alz$tau, breaks = 30)
hist(Alz$age, breaks = 30)
```

**Scaling Data**
Using the **scale()** function we can get all of our variables to have a mean of 0 and standard deviation of 1.

```{r}
tau_scaled <- scale(Alz$tau, center = TRUE)
age_scaled <- scale(Alz$age, center = TRUE)
hist(tau_scaled, breaks = 30)
hist(age_scaled, breaks = 30)
```

**Dummy variables**
Some machine learning algorithms can't handle factor variables.  They must be turned into numeric type data.  This process is referred to as creating dummy variables

So which of our variables are factor variables?

```{r}
Alz_var_classes <- sapply(Alz, class)
which(Alz_var_classes == 'factor')
```

Genotype, response and gender are all factor variables.  


Let's see how we can turn one of them into a dummy variable using **dummyVars()** in caret

```{r}
dmy <- dummyVars(~ Genotype, data = Alz)
dmy_df <- data.frame(predict(dmy, newdata = Alz))
head(dmy_df)
head(Alz$Genotype)
```
